# Multivariate Imputation by Chained Equations (MICE)

## The `mice` package  

[Package Website link](https://amices.org/mice/)

> The mice package implements a method to deal with missing data. The package creates multiple imputations (replacement values) for multivariate missing data. The method is based on **Fully Conditional Specification**, where each incomplete variable is imputed by a separate model. The `mice` algorithm can impute mixes of continuous, binary, unordered categorical and ordered categorical data. In addition, `mice` can impute continuous two-level data, and maintain consistency between imputations by means of passive imputation. Many diagnostic plots are implemented to inspect the quality of the imputations.

[Winnipeg workshop: Handling missing data in `R` with `mice`](https://amices.org/Winnipeg/)


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(mice)
```


```{r, eval=FALSE}
?nhanes
```



```{r}
nhanes
```


```{r}
tibble::glimpse(nhanes)
```



```{r}
summary(nhanes)
```


```{r}
df_nhanes <- nhanes %>% 
  dplyr::mutate(age = factor(age,
                             levels = c(1, 2,3),
                             labels = c("20-39", "40-59", "60+"))) %>% 
  dplyr::mutate(hyp = factor(hyp,
                             levels = c(1, 2),
                             labels = c("no", "yes")))
```

```{r}
summary(df_nhanes)
```



```{r}
mice::md.pattern(df_nhanes)
```

## multiple impute the missing values

```{r}
imp <- mice::mice(data = df_nhanes,   #name of dataset
                  m = 4,              # number of copies
                  seed = 432)
```

```{r}
imp
```



```{r}
mice::stripplot(imp, chl,
                pch = 19,
                xlab = "Imputation number")
```



```{r}
fit_lm_orig <- lm(chl ~ age + bmi + hyp,
                  data = df_nhanes)
```

```{r}
summary(fit_lm_orig)
```

## Run the model on each imputate datset

```{r}
fit_lm_imp <- with(imp, lm(chl ~ age + bmi + hyp))
```


```{r}
fit_lm_imp
```

## pool the results into one set of results

```{r}
fit_lm_pooled <-mice::pool(fit_lm_imp)
```


```{r}
summary(fit_lm_pooled)
```

